---
title: "Another R Tutorial"
author: "steven"
date: "October 14, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This introduction will focus on base R (and introduce tidyverse and RDataTable at the end).  
We will begin with creating vectors.
```{r vectors}
x1 <- rpois(10, lambda = 1) 

length(x1)

x1 

x1 + 1 

x1 ^ 2 
```
```{r}
x2 <- rexp(20)
x2
```

Let's do a quick plot of `r x1`. If you only supply 1 vector, the `r plot` function will plot by index.
```{r}
plot(x1)
```

```{r}
plot(runif(10), x1)
```


Vectors allow index retrieval. "letters" is a built-in R vector that contains the lowercase alphabet.  
Index starts at 1.
```{r vectors2}
x3 <- letters[10]
x3
```

Lists are the foundation of R's various data structures. They're powerful but can be tricky to use at first.  
Each element of the list can be a different data structure. Here, we'll just use the three vectors from before.
```{r lists}
list1 <- list(x1, x2, x3) 
```


Base R has a famous family of functions that I like to call the "apply" family. `r lapply` applies a function to each element of the list. Below, we'll drop the last element from the list.
```{r}
lapply(list1[[-3]], mean) 

lapply(list1[[-3]], function(x) sum(x^2)) 
```

You will most likely use dataframes instead. Dataframes can be thought of as a special case of lists (i.e. a "flat" list). We'll read in a CSV file.
```{r}
pokemon <- read.csv("pokemon.csv", header = T, stringsAsFactors = F)
head(pokemon)
```

Let's take a quick look at what columns are in the dataframe.
```{r}
colnames(pokemon)
```

One recurring theme in R is that there are several ways to do the same thing. Here's one way of subsetting columns.
```{r}
pokemon2 <- pokemon[, c("type1", "name", "is_legendary", "generation", "evolution_stage", "attack", "defense", "speed", "sp_attack", "sp_defense")] 
```

You can also use column indices.
```{r}
head(pokemon[, c(37, 31, 41)])
```

Even logical indices!
```{r}
head(pokemon[, c(T, T, F, T)])
```

We can use the `r subset` function to, well, subset. We'll subset the dataframe to the first 2 generations. We'll then use the `r unique` function as a check.
```{r}
pokemon_first_2_gen <- subset(pokemon2, pokemon2$generation %in% c(1, 2)) 

unique(pokemon_first_2_gen$generation)
```

If you look at the documentation, there are many ways to set up the `r aggregate` function. Here's what I think is the easiest. We'll find the average stats by type and generation.
```{r}
pokemon_stats <- aggregate(cbind(attack, defense, speed, sp_attack, sp_defense) ~ type1 + generation, data = pokemon_first_2_gen, mean)
pokemon_stats
```

Let's do a boxplot. we'll subset to just the first generation as to not crowd the plot.
```{r}
boxplot(attack ~ type1 + generation, data = subset(pokemon2, pokemon2$generation == 1)) 
```

Let's plot the aggregated data. We'll use `r ggplot2` as it will give us more control over the plot.
```{r}
library(ggplot2)
library(magrittr)

pokemon_stats %>%
  mutate(type = as.factor(type1)) %>%
ggplot(aes(type, attack, color = generation)) + 
  geom_point(stat = "identity") +  
  scale_fill_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r dplyr}
library(dplyr)
pkmn_stats_tidy <- pokemon_first_2_gen %>% 
  filter(generation == 1) %>% 
  group_by(type1) %>% 
  summarize(attack = mean(attack), 
            defense = mean(defense), 
            speed = mean(speed), 
            sp_attack = mean(sp_attack), 
            sp_defense = mean(sp_defense))
pkmn_stats_tidy
```

```{r}
print.data.frame(pkmn_stats_tidy)
```

```{r data.table}
library(data.table) 

pokemon_dt <- fread(file = "pokemon.csv") 
pkmn_stats_dt <- pokemon_dt[generation == 1,  
                             .(attack = mean(attack), 
                                defense = mean(defense), 
                                speed = mean(speed), 
                                sp_attack = mean(sp_attack), 
                                sp_defense = mean(sp_defense)), 
                             by = "type1"][order(type1)] 

pkmn_stats_dt 
```

